---
title: "Complete Schedule Coverage"
author: "PAHO/CIM"
date: "April 11, 2025"
output: 
  html_document:
    self_contained: true
    theme: cosmo
    highlight: tango
    code_folding: show
    code_download: true
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(pahoabc)
library(dplyr)
library(knitr)
```

# Rationale

The purpose of this analysis is to evaluate whether the target population has been fully vaccinated according to the country's defined immunization schedule for a specific location and cohort. To construct this indicator, it is essential to assess if each individual received all doses specified in the schedule. For instance, a country's complete immunization schedule might include: 1 dose of BCG for newborns, 3 doses of DTP-containing vaccine, 1 dose of MMR, and 1 dose of yellow fever vaccine. Unlike administrative records, which often use a tracer dose from the end of the schedule (e.g., DTP3), the use of Electronic Immunization Records (EIR) allows for the evaluation of all vaccines within the schedule.

> **Note**
> 
> All functions used for complete schedule coverage analyses within PAHOabc begin with the `cs_` prefix.

# Country Vaccination Schedule

To enable this package to analyze the vaccination schedule, it is necessary to provide the country's immunization schedule using the same format as the example schedule shown below.

```{r schedule=example, message=FALSE}

# show example vaccination schedule
pahoabc::pahoabc.schedule %>% 
  kable(
    caption = "Example country vaccination schedule",
    col.names = c("Dose", "Recommended age (in years)")
  )
```

# Complete Schedule Coverage Computation

## Expected Workflow

The `cs_coverage()`, `cs_barplot()` and `cs_upset()` functions provide a way of analyzing and visualizing complete schedule coverage. Their relationships are shown in Figure 1.

![Figure 1. Expected workflow for complete schedule coverage analysis.](cs_workflow.svg)

## Example

Below is a simplified example of the calculation of complete schedule coverage for the first subnational administrative level, focusing on the birth cohort of the year 2022.

```{r cov-analysis-example, message=FALSE}

# compute coverage
coverage_df <- cs_coverage(
  data.EIR = pahoabc.EIR,
  data.schedule = pahoabc.schedule,
  geo_level = "ADM1",
  birth_cohorts = 2022 # this can be a vector of years
)

# show results in table
coverage_df %>%
  kable(digits = 2, caption = "Complete schedule coverage by ADM1 level")
```

## Bar plot

It is possible to display the previously obtained results in a bar plot with `cs_barplot()`.

```{r cov-viz-example, message=FALSE, fig.width=12, fig.height=6}

# visualize output from cs_coverage()
coverage_plot <- cs_barplot(data = coverage_df)

# show
coverage_plot
```

## Upset plot

The `cs_upset()` function allows us to produce an upset plot to evaluate the percentage of individuals included in the EIR who have received any of the possible combinations of the available vaccines. The orange bar represents the proportion of individuals who have received all the evaluated vaccines, thus indicating the complete schedule coverage. The horizontal bars show the proportion of individuals who received each specific vaccine (i.e., the coverage for that vaccine).

```{r cov-upset-example, message=FALSE, fig.width=12, fig.height=6}

# produce upset plot
coverage_upset <- cs_upset(
  data.EIR = pahoabc.EIR,
  data.schedule = pahoabc.schedule,
  birth_cohort = 2022
)

# show
coverage_upset
```

In the example the complete schedule coverage reached 13.7% and is represented by the orange bar. 
It is noteworthy that 24.1% of individuals in the EIR received only the BCG vaccine, and that the third most frequent combination (9.1%) consists of individuals who received all other vaccines except BCG. This could indicate an error in recording a newborn's BCG vaccine using temporary identification numbers.
