---
title: "Complete Schedule Coverage"
author: "PAHO/CIM"
date: "April 11, 2025"
output: 
  html_document:
    self_contained: true
    theme: cosmo
    highlight: tango
    code_folding: show
    code_download: true
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(pahoabc)
library(dplyr)
library(knitr)
```

# Rationale

The purpose of this analysis is to evaluate whether the target population has been fully vaccinated according to the country's defined immunization schedule for a specific location and cohort. To construct this indicator, it is essential to assess if each individual received all doses specified in the schedule. For instance, a country's complete immunization schedule might include: 1 dose of BCG for newborns, 3 doses of DTP-containing vaccine, 1 dose of MMR, and 1 dose of yellow fever vaccine. Unlike administrative records, which often use a tracer dose from the end of the schedule (e.g., DTP3), the use of Electronic Immunization Records (EIR) allows for the evaluation of all vaccines within the schedule.

> **Note**
> 
> All functions used for complete schedule coverage analyses within PAHOabc begin with the `cs_` prefix.

# Glossary

## Political Geographic Boundaries

Throughout this vignette, we will make reference to different levels of political geographic boundaries. These levels are defined by the country/territory and can have different names from country to country. PAHOabc is agnostic to these names and requires the user to recode their variable names to fit the PAHOabc structure.

Namely, PAHOabc can distinguish three administrative levels, which listed from top to bottom level are: ADM0, ADM1 and ADM2.

1. ADM0: The country. This is the top-most administrative level.
2. ADM1: The first geographic subdivision in the country. ADM0 contains several ADM1 subdivisions.
3. ADM2: The second geographic subdivision in the country. Each ADM1 contains several ADM2 subdivisions.

## Vaccine Names

Our example datasets use a specific naming convention for the vaccine doses. This might not match your own naming schemes, but it will not affect the usage of the package as long as you make sure to be consistent when naming vaccine doses.

For example, the vaccine names used in the PAHOabc package are:

| Abbreviation | Full Name                                                        |
|--------------|------------------------------------------------------------------|
| SRP1         | Measles, Rubella, and Mumps Vaccine (1st dose)                   |
| DTP1         | Diphtheria, Tetanus, and Pertussis Vaccine (1st dose)            |
| DTP2         | Diphtheria, Tetanus, and Pertussis Vaccine (2nd dose)            |
| DTP3         | Diphtheria, Tetanus, and Pertussis Vaccine (3rd dose)            |
| BCG RN       | Bacillus Calmette–Guérin Vaccine (at birth)                      |
| YFV1         | Yellow Fever Vaccine (1st dose)                                  |

# Usage

## Install Package

The first step to run the residence occurrence analyses is to install the PAHOabc package available on GitHub.

```{r, install_package, eval=FALSE}
devtools::install_github("IM-Data-PAHO/pahoabc")
```

## Load Data

The functions in this module require you to provide three datasets in a specific format.

1. Your EIR.
2. The vaccination schedule related to this EIR.
3. A table with population denominators depending on the geographic level of your analysis.
    1. This is optional for the `cs_coverage()` function (but not shown in this vignette).

To make it easy for you to test out PAHOabc's functionality and understand the structure we require for your datasets, we will now explore the example datasets provided by PAHOabc that are required by this module.

### EIR

The `pahoabc::pahoabc.EIR` data frame provides a simulated, nominal-level table representing individual vaccination events from an Electronic Immunization Registry (EIR). Each row corresponds to a single vaccination act for a person, including information on their residence, where they were vaccinated (occurrence), their date of birth, and the vaccine dose received.

```{r pahoabc-EIR, message=FALSE}
pahoabc.EIR %>% head() %>% kable(caption = "Example Electronic Immunization Registry")
```

- `ID`: Unique person identification number.
- `date_birth`: Date of birth of person. 
- `date_vax`: Date of vaccination event. 
- ADM1: Refers to the first geographic administrative level of the country.
- ADM2: Refers to the second geographic administrative level of the country.
- Residence: Refers to the place where the person lives.  
- Occurrence: Refers to the place where the vaccination event occurred.  
- `dose`: A combined variable representing the vaccine type and its corresponding dose number. For example, DTP1 refers to the first dose of a vaccine containing diphtheria, tetanus, and pertussis components.

### Vaccination Schedule

The `pahoabc::pahoabc.schedule` dataset defines the national immunization schedule, listing each vaccine dose and its recommended age of administration (in months). It helps the package determine whether an individual is up to date with their vaccinations.

```{r pahoabc-schedule, message=FALSE}
pahoabc.schedule %>% kable(caption = "Example Immunization Schedule")
```

- `dose`: A combined variable representing the vaccine type and its corresponding dose number. For example, DTP1 refers to the first dose of a vaccine containing diphtheria, tetanus, and pertussis components.
- `age_schedule`: The recommended age of administration of the corresponding `dose` in months.

> **Note**
> 
> The `dose` names must match exactly those in the `dose` column of the `pahoabc.EIR` dataset.

### Population Denominators

This module will require you to provide a table with population denominators. The geographic level of this table will be directly dependent on the geographic level of your analysis. For example, if you were to perform an analysis at ADM1 level, the required population denominator table must contain the following.

```{r pahoabc-pop-ADM1, message=FALSE}
pahoabc.pop.ADM1 %>% head() %>% kable(caption = "Example Population at ADM1")
```

Note that this table is in long format and contains the corresponding `population` of children of a specific `age` during a given `year` for a certain `ADM1` level.

## Complete Schedule Coverage Analysis

### Expected Workflow

The `cs_coverage()`, `cs_barplot()` and `cs_upset()` functions provide a way of analyzing and visualizing complete schedule coverage. Their relationships are shown in Figure 1.

![Figure 1. Expected workflow for complete schedule coverage analysis.](cs_workflow.svg)

<br>

### Example

Below is a simplified example of the calculation of complete schedule coverage for the first subnational administrative level, focusing on the birth cohort of the year 2022.

```{r cov-analysis-example, message=FALSE}

# compute coverage
coverage_df <- cs_coverage(
  data.EIR = pahoabc.EIR,
  data.schedule = pahoabc.schedule,
  geo_level = "ADM1",
  birth_cohorts = 2022 # this can be a vector of years
)

# show results in table
coverage_df %>%
  kable(digits = 2, caption = "Complete schedule coverage by ADM1 level")
```

### Bar plot

It is possible to display the previously obtained results in a bar plot with `cs_barplot()`.

```{r cov-viz-example, message=FALSE, fig.width=12, fig.height=6}

# visualize output from cs_coverage()
coverage_plot <- cs_barplot(data = coverage_df)

# show
coverage_plot
```

### Upset plot

The `cs_upset()` function allows us to produce an upset plot to evaluate the percentage of individuals included in the EIR who have received any of the possible combinations of the available vaccines. The orange bar represents the proportion of individuals who have received all the evaluated vaccines, thus indicating the complete schedule coverage. The horizontal bars show the proportion of individuals who received each specific vaccine (i.e., the coverage for that vaccine).

```{r cov-upset-example, message=FALSE, fig.width=12, fig.height=6}

# produce upset plot
coverage_upset <- cs_upset(
  data.EIR = pahoabc.EIR,
  data.schedule = pahoabc.schedule,
  birth_cohort = 2022
)

# show
coverage_upset
```

In the example the complete schedule coverage reached 13.7% and is represented by the orange bar.

It is noteworthy that 24.1% of individuals in the EIR received only the BCG vaccine, and that the third most frequent combination (9.1%) consists of individuals who received all other vaccines except BCG. This could indicate an error in recording a newborn's BCG vaccine using temporary identification numbers.
