---
title: "Date Consistency Comparison"
author: "PAHO/CIM"
date: "June 1, 2026"
output:
  html_document:
    self_contained: true
    theme: cosmo
    highlight: tango
    code_folding: show
    code_download: true
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(pahoabc)
library(dplyr)
library(knitr)
```

# Rationale

Electronic Immunization Registries (EIRs) often contain date fields that are logically constrained relative to one another. A common example is the relationship between a vaccination date and a date of birth: a vaccination event cannot occur before a person is born. When such violations exist in the data, they indicate data entry errors or record linkage issues that can bias downstream analyses.

The date consistency comparison module allows users to systematically identify, quantify, and visualize these inconsistencies between any two date variables in the EIR.

> **Note**
>
> All functions used for date consistency comparison analyses within PAHOabc begin with the `dcc_` prefix.

# Glossary

## Political Geographic Boundaries

PAHOabc can distinguish three administrative levels, listed from top to bottom:

1. **ADM0**: The country — the top-most administrative level.
2. **ADM1**: The first geographic subdivision within the country.
3. **ADM2**: The second geographic subdivision. Each ADM1 contains several ADM2 subdivisions.

## Consistency Categories

| Category | Meaning |
|---|---|
| `CONSISTENT` | `date_2` ≤ `date_1` — the chronological order is valid. |
| `INCONSISTENT` | `date_2` > `date_1` — the chronological order is violated. |
| `DATE_MISSING` | At least one of the two dates is `NA`. |

# Usage

## Install Package

```{r install_package, eval=FALSE}
devtools::install_github("IM-Data-PAHO/pahoabc")
```

## Load Data

The functions in this module require your **EIR dataset** in the standard PAHOabc format.

### EIR

The `pahoabc::pahoabc.EIR` data frame provides a simulated, nominal-level table representing individual vaccination events from an Electronic Immunization Registry (EIR). Each row corresponds to a single vaccination act for a person.

```{r pahoabc-EIR, message=FALSE}
pahoabc.EIR %>% head() %>% kable(caption = "Example Electronic Immunization Registry")
```

- `ID`: Unique person identification number.
- `date_birth`: Date of birth of person.
- `date_vax`: Date of vaccination event.
- `ADM1_residence` / `ADM2_residence`: First and second geographic administrative levels of residence.
- `dose`: Combined variable representing vaccine type and dose number (e.g., DTP1).

## Date Consistency Comparison Analysis

### Expected Workflow

The `dcc_` suite contains four functions that work together:

1. `dcc_rate()` — computes consistency rates aggregated by geographic level.
2. `dcc_barplot()` — visualizes the output of `dcc_rate()` as a stacked bar chart.
3. `dcc_inconsistent()` — lists individual inconsistent and missing records with their day difference.
4. `dcc_inconsistent_plot()` — visualizes the distribution of day differences for inconsistent records.

### Step 1 — Compute consistency rates with `dcc_rate()`

`dcc_rate()` compares two date variables in the EIR and returns, for each geographic unit, the count and percentage of records that are `CONSISTENT`, `INCONSISTENT`, or `DATE_MISSING`.

Here we check whether `date_vax` (date_1, the later date) is consistent with `date_birth` (date_2, the earlier date) at the ADM1 level.

```{r dcc-rate, message=FALSE}
rate_df <- dcc_rate(
  data.EIR = pahoabc.EIR,
  date_1   = "date_vax",
  date_2   = "date_birth",
  geo_level = "ADM1"
)

rate_df %>% kable(digits = 2, caption = "Date consistency rates by ADM1")
```

#### Parameters accepted

- `data.EIR`: EIR data frame in PAHOabc format.
- `date_1`: Name of the date variable being checked (the chronologically later date).
- `date_2`: Name of the date variable checked against (the chronologically earlier date).
- `geo_level`: `"ADM0"` (default), `"ADM1"`, or `"ADM2"`.

### Step 2 — Visualize rates with `dcc_barplot()`

The output of `dcc_rate()` can be passed directly to `dcc_barplot()` to produce a stacked bar chart showing the breakdown of consistency categories per geographic unit.

```{r dcc-barplot, message=FALSE, fig.width=10, fig.height=5}
dcc_barplot(
  data        = rate_df,
  date_1_name = "Vaccination date",
  date_2_name = "Date of birth"
)
```

#### Parameters accepted

- `data`: Output of `dcc_rate()`.
- `date_1_name`: Display label for `date_1` in the plot subtitle. Default: `"date 1"`.
- `date_2_name`: Display label for `date_2` in the plot subtitle. Default: `"date 2"`.
- `within_ADM1`: Character vector to filter to specific ADM1 units when `geo_level = "ADM2"`. Default: `NULL`.
- `plot_missing`: Boolean. If `TRUE` (default), `DATE_MISSING` records are shown as a separate bar segment so all bars sum to 100%.

#### Interpretation

Each bar represents a geographic unit and is divided into three segments: `CONSISTENT` (blue), `INCONSISTENT` (orange), and `DATE_MISSING` (grey). Bars that are predominantly orange or grey indicate units with data quality issues requiring attention.

### Step 3 — Inspect individual inconsistent records with `dcc_inconsistent()`

`dcc_inconsistent()` returns a record-level table of all `INCONSISTENT` and `DATE_MISSING` entries, including the day difference between the two dates.

```{r dcc-inconsistent, message=FALSE}
inconsistent_df <- dcc_inconsistent(
  data.EIR = pahoabc.EIR,
  date_1   = "date_vax",
  date_2   = "date_birth"
)

inconsistent_df %>% head(10) %>% kable(caption = "Sample of inconsistent and missing records")
```

#### Parameters accepted

- `data.EIR`: EIR data frame in PAHOabc format.
- `date_1`: Name of the date variable being checked.
- `date_2`: Name of the date variable checked against.

The `diff` column reports `date_1 - date_2` in days. Negative values indicate `date_1` precedes `date_2` (i.e., the inconsistency). This table can be exported for record-level correction workflows.

### Step 4 — Visualize the distribution of day differences with `dcc_inconsistent_plot()`

`dcc_inconsistent_plot()` produces a histogram of the **absolute** day differences for inconsistent records, grouped into customizable bins. This helps characterize the severity of inconsistencies: small differences may reflect data entry typos, while large differences suggest more systematic issues.

```{r dcc-inconsistent-plot, message=FALSE, fig.width=10, fig.height=5}
dcc_inconsistent_plot(
  data.EIR    = pahoabc.EIR,
  date_1      = "date_vax",
  date_2      = "date_birth",
  date_1_name = "Vaccination date",
  date_2_name = "Date of birth"
)
```

#### Parameters accepted

- `data.EIR`: EIR data frame in PAHOabc format.
- `date_1`: Name of the date variable being checked.
- `date_2`: Name of the date variable checked against.
- `date_1_name`: Display label for `date_1` in the plot subtitle. Default: `"date 1"`.
- `date_2_name`: Display label for `date_2` in the plot subtitle. Default: `"date 2"`.
- `day_groups`: List of `c(lower, upper)` numeric pairs defining the histogram bins. The last bin is always extended to `Inf` and labelled `"X+"`. Default: 10-day bins from 0 to 100.

To use custom bin widths — for example, finer resolution for small differences:

```{r dcc-inconsistent-plot-custom, message=FALSE, fig.width=10, fig.height=5}
dcc_inconsistent_plot(
  data.EIR    = pahoabc.EIR,
  date_1      = "date_vax",
  date_2      = "date_birth",
  date_1_name = "Vaccination date",
  date_2_name = "Date of birth",
  day_groups  = list(c(0, 1), c(1, 7), c(7, 30), c(30, 90), c(90, 180))
)
```

#### Interpretation

The x-axis shows the absolute number of days between the two dates, regardless of direction (i.e., whether `date_1` precedes `date_2` or vice versa). Records concentrated in the first bins (0–9 days) are likely simple data entry errors, while records in the 100+ category may indicate systematic record linkage problems.

# Summary

The `dcc_` module provides a complete pipeline for assessing date consistency in EIR data. Starting from `dcc_rate()` for a geographic overview, through `dcc_barplot()` for visualization, to `dcc_inconsistent()` for record-level inspection and `dcc_inconsistent_plot()` for characterizing the magnitude of errors, the suite equips data managers and public health analysts with the tools needed to detect, quantify, and prioritize data quality corrections before any downstream analysis.
