---
title: "Understanding the Example Datasets in PAHOabc"
author: "PAHO/CIM"
date: "April 14, 2025"
output: 
  html_document:
    self_contained: true
    theme: cosmo
    highlight: tango
    code_folding: show
    code_download: true
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(pahoabc)
library(dplyr)
library(knitr)
```

# Overview

The PAHOabc package includes a set of example datasets to help users understand how to use its functions effectively. These datasets simulate a fictional country's immunization system and are structured to work seamlessly with the tools provided by the package. This vignette provides an overview of each dataset and describes how they are used throughout the package's modules.

# List of example datasets

## pahoabc.EIR

The `pahoabc::pahoabc.EIR` dataframe is the most important dataset in the package. It is a nominal-level table representing individual vaccination events from an Electronic Immunization Registry (EIR). Each row corresponds to a single vaccination act for a person, including information on their residence, where they were vaccinated, their date of birth, and the vaccine dose received.

```{r pahoabc-EIR, message=FALSE}
pahoabc.EIR %>% head() %>% kable(caption = "Example Electronic Immunization Registry")
```

All records include ADM1 and ADM2 identifiers for both residence and occurrence, making it possible to analyze population flows and coverage across different geographic levels.

## pahoabc.schedule

The `pahoabc::pahoabc.schedule` dataset defines the national immunization schedule, listing each vaccine dose and its recommended age of administration (in years). It helps the package determine whether an individual is up to date with their vaccinations.

```{r pahoabc-schedule, message=FALSE}
pahoabc.schedule %>% kable(caption = "Example Immunization Schedule")
```

> **Note**
> 
> The `dose` names must match exactly those in the `dose` column of the `pahoabc.EIR` dataset.

## Population datasets

Population datasets provide population estimates at different geographic levels, for different years and children of varying ages. Depending on the geographic level at which we want to perform an analysis, PAHOabc may require you to provide the corresponding population to use as a denominator (mainly for coverage analyses).

### pahoabc.pop.ADM0

The `pahoabc::pahoabc.pop.ADM0` dataset provides aggregated population estimates at the national level (ADM0).

```{r pahoabc-pop-ADM0, message=FALSE}
pahoabc.pop.ADM0 %>% kable(caption = "Example Population at ADM0")
```

### pahoabc.pop.ADM1

The `pahoabc::pahoabc.pop.ADM1` dataset provides population data aggregated at the first administrative level (ADM1), typically corresponding to regions, provinces, or states.

```{r pahoabc-pop-ADM1, message=FALSE}
pahoabc.pop.ADM1 %>% head() %>% kable(caption = "Example Population at ADM1")
```

### pahoabc.pop.ADM2

The `pahoabc::pahoabc.pop.ADM2` dataset provides more granular population data by both ADM1 and ADM2 levels, allowing analyses at a finer geographic resolution.

```{r pahoabc-pop-ADM2, message=FALSE}
pahoabc.pop.ADM2 %>% head() %>% kable(caption = "Example Population at ADM2")
```

# Summary

All example datasets are loaded automatically with the package. You can explore their structure using `glimpse()` or `View()`. They are designed to work seamlessly with the default arguments in all functions, allowing you to test and understand each module with minimal setup.

Here is a summary of the available datasets for your reference.

| Dataset             | Description                                                   |
|---------------------|---------------------------------------------------------------|
| `pahoabc.EIR`       | Electronic Immunization Registry                              |
| `pahoabc.schedule`  | Vaccination schedule with dose names and ages                 |
| `pahoabc.pop.ADM0`  | Aggregated national-level population estimates                |
| `pahoabc.pop.ADM1`  | ADM1-level population estimates                               |
| `pahoabc.pop.ADM2`  | ADM2-level population estimates                               |
