---
title: "Using our Example Datasets"
author: "PAHO/CIM"
date: "April 14, 2025"
output: 
  html_document:
    self_contained: true
    theme: cosmo
    highlight: tango
    code_folding: show
    code_download: true
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
library(pahoabc)
library(dplyr)
library(knitr)
```

# Rationale

The PAHOabc package includes a set of example datasets to help users understand how to use its functions effectively. These datasets simulate a fictional country's immunization information system and are structured to work seamlessly with the tools provided by the package. This vignette provides an overview of each dataset and describes how they are used throughout the package's modules.

# Glossary

## Political Geographic Boundaries

Throughout this vignette, we will make reference to different levels of political geographic boundaries. These levels are defined by the country/territory and can have different names from country to country. PAHOabc is agnostic to these names and requires the user to recode their variable names to fit the PAHOabc structure.

Namely, PAHOabc can distinguish three administrative levels, which listed from top to bottom level are: ADM0, ADM1 and ADM2.

1. ADM0: The country. This is the top-most administrative level.
2. ADM1: The first geographic subdivision in the country. ADM0 contains several ADM1 subdivisions.
3. ADM2: The second geographic subdivision in the country. Each ADM1 contains several ADM2 subdivisions.

![](ADM_explanation.svg)

## Vaccine Names

Our example datasets use a specific naming convention for the vaccine doses. This might not match your own naming schemes, but it will not affect the usage of the package as long as you make sure to be consistent when naming vaccine doses.

For example, the vaccine names used in the PAHOabc package are:

| Abbreviation | Full Name                                                        |
|--------------|------------------------------------------------------------------|
| SRP1         | Measles, Rubella, and Mumps Vaccine (1st dose)                   |
| DTP1         | Diphtheria, Tetanus, and Pertussis Vaccine (1st dose)            |
| DTP2         | Diphtheria, Tetanus, and Pertussis Vaccine (2nd dose)            |
| DTP3         | Diphtheria, Tetanus, and Pertussis Vaccine (3rd dose)            |
| BCG RN       | Bacillus Calmette–Guérin Vaccine (at birth)                      |
| YFV1         | Yellow Fever Vaccine (1st dose)                                  |

# List of example datasets

## pahoabc.EIR

The `pahoabc::pahoabc.EIR` dataframe is the most important dataset in the package. It is a nominal-level table representing individual vaccination events from an Electronic Immunization Registry (EIR). Each row corresponds to a single vaccination act for a person, including information on their residence, where they were vaccinated, their date of birth, and the vaccine dose received.

```{r pahoabc-EIR, message=FALSE}
pahoabc.EIR %>% head() %>% kable(caption = "Example Electronic Immunization Registry")
```

- `ID`: Unique person identification number.
- `date_birth`: Date of birth of person. 
- `date_vax`: Date of vaccination event. 
- ADM1: Refers to the first geographic administrative level of the country.
- ADM2: Refers to the second geographic administrative level of the country.
- Residence: Refers to the place where the person lives.  
- Occurrence: Refers to the place where the vaccination event occurred.  
- `dose`: A combined variable representing the vaccine type and its corresponding dose number. For example, DTP1 refers to the first dose of a vaccine containing diphtheria, tetanus, and pertussis components.

## pahoabc.schedule

The `pahoabc::pahoabc.schedule` dataset defines the national immunization schedule, listing each vaccine dose and its recommended age of administration (in months). It helps the package determine whether an individual is up to date with their vaccinations.

```{r pahoabc-schedule, message=FALSE}
pahoabc.schedule %>% kable(caption = "Example Immunization Schedule")
```

- `dose`: A combined variable representing the vaccine type and its corresponding dose number. For example, DTP1 refers to the first dose of a vaccine containing diphtheria, tetanus, and pertussis components.
- `age_schedule`: The recommended age of administration of the corresponding `dose` in months.

> **Note**
> 
> The `dose` names must match exactly those in the `dose` column of the `pahoabc.EIR` dataset.

## Population Datasets

Population datasets provide population estimates at different geographic levels, for various years and children of varying ages. Depending on the geographic level at which we want to perform an analysis, PAHOabc may require you to provide the corresponding population to use as a denominator (mainly for coverage analyses).

### pahoabc.pop.ADM0

The `pahoabc::pahoabc.pop.ADM0` dataset provides aggregated population estimates at the national level (ADM0).

```{r pahoabc-pop-ADM0, message=FALSE}
pahoabc.pop.ADM0 %>% kable(caption = "Example Population at ADM0")
```

This table is in long format and contains the corresponding `population` of children of a specific `age` during a given `year`.

### pahoabc.pop.ADM1

The `pahoabc::pahoabc.pop.ADM1` dataset provides population data aggregated at the first subnational administrative level (ADM1), typically corresponding to regions, provinces, or states.

```{r pahoabc-pop-ADM1, message=FALSE}
pahoabc.pop.ADM1 %>% head() %>% kable(caption = "Example Population at ADM1")
```

Note that this table is in long format and contains the corresponding `population` of children of a specific `age` during a given `year` at the `ADM1` level.

### pahoabc.pop.ADM2

The `pahoabc::pahoabc.pop.ADM2` dataset provides more granular population data by both ADM1 and ADM2 levels, allowing analyses at a finer geographic resolution.

```{r pahoabc-pop-ADM2, message=FALSE}
pahoabc.pop.ADM2 %>% head() %>% kable(caption = "Example Population at ADM2")
```

Note that this table is in long format and contains the corresponding `population` of children of a specific `age` during a given `year` at the `ADM2` level.

# Summary

All example datasets are loaded automatically with the package. You can explore their structure using `glimpse()` or `View()`. They are designed to work seamlessly with the default arguments in all functions, allowing you to test and understand each module with minimal setup.

Here is a summary of the available datasets for your reference.

| Dataset             | Description                                                   |
|---------------------|---------------------------------------------------------------|
| `pahoabc.EIR`       | Electronic Immunization Registry                              |
| `pahoabc.schedule`  | Vaccination schedule with dose names and ages                 |
| `pahoabc.pop.ADM0`  | Aggregated national-level population estimates                |
| `pahoabc.pop.ADM1`  | ADM1-level population estimates                               |
| `pahoabc.pop.ADM2`  | ADM2-level population estimates                               |
